---
title: "Reinforcement learning with rewards you can check"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reinforcement learning with rewards you can check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`dragon_reinforce()` trains a model with Group Relative Policy Optimization
(GRPO), the method behind the recent reasoning models, in its plain on-policy
form. For every prompt the model writes several answers, each is scored by
rewards you define, and the model is nudged towards the answers that beat
their group's average. A KL penalty against the model it started from keeps
it from drifting.

## When it helps, and when it does not

Reinforcement learning on a small model works when the reward is
**verifiable**: something a function can check without opinion.

* A number that must match: arithmetic, unit conversion, extracted totals.
* A format that must hold: valid JSON with given keys, a ticket id at the
  start, a required section.
* A budget: at most 300 characters, at least three bullet points.
* A test that passes.

It works poorly as a substitute for preference data on goals such as "be
more helpful" or "sound friendlier". A reward that is really a model's
opinion is noisy, and a small policy will find the noise. For those goals
use `dragon_prefer()` with judge-ranked pairs instead; see the post-training
loop vignette.

## Data: prompts and references

Rows carry a prompt and, optionally, a reference answer for rewards that
compare against one. Every other column travels along as `fields` for custom
rewards.

```{r}
library(dragonfarm)

a <- sample(10:999, 400, replace = TRUE)
b <- sample(10:999, 400, replace = TRUE)
math <- data.frame(
  question = sprintf("What is %d + %d? Reply with just the number.", a, b),
  answer = as.character(a + b)
)
prompts <- dragon_map_prompts(dragon_dataset(math), prompt = "question", reference = "answer")
dragon_preview(prompts, n = 1)
```

## Rewards

`dragon_reward()` builds one reward; a list of them is summed by weight.

```{r}
rewards <- list(
  dragon_reward("numeric"),                            # last number equals the reference
  dragon_reward("length", max_chars = 12, weight = 0.3) # keep it terse
)
```

The built-ins: `"exact"`, `"contains"`, `"numeric"`, `"regex"` (with
`pattern`), `"json"` (with optional `keys`, partial credit per key),
`"length"` (with `min_chars` and `max_chars`), and `"keyword"` (with `words`
and `mode`). For anything else, `"custom"` points at a Python file:

```{r}
writeLines(c(
  "def reward(prompt, completion, reference, row):",
  "    # 1.0 when the reply mentions the product named in the row",
  "    return 1.0 if row.get('product', '') and row['product'] in completion else 0.0"
), "product_reward.py")
mention <- dragon_reward("custom", file = "product_reward.py", name = "mentions_product")
```

The file is copied into the run, so the run stays self-contained and travels
with the cloud bundle.

## Train

Start from a fine-tuned run when you have one; the RL stage folds its
adapters in before adding its own.

```{r}
rl <- dragon_reinforce(
  prompts, sft,                    # or a model id such as "Qwen/Qwen2.5-0.5B-Instruct"
  rewards = rewards,
  group_size = 6,                  # answers sampled per prompt
  beta = 0.04,                     # KL penalty towards the starting model
  temperature = 1.0,               # sampling temperature; must be above 0
  max_new_tokens = 16,
  args = dragon_train_args(learning_rate = 1e-5, epochs = 1, batch_size = 4, grad_accum = 1, save_steps = 20),
  wait = TRUE
)
```

A step samples `batch_size * group_size` completions, scores them, and takes
one optimizer step. Sampling dominates the cost, so short `max_new_tokens`
and modest `group_size` keep steps fast. Progress rows carry `reward`,
`reward_std`, `kl`, and `completion_len`, plus one column per reward:

```{r}
dragon_progress(rl)[, c("step", "reward", "reward_numeric", "reward_length", "kl", "completion_len")]
```

The Monitor panel plots mean reward on a second axis next to the loss.

## Read the result

```{r}
dragon_evaluate(rl)   # mean held-out reward, per reward
dragon_generate(rl, "What is 417 + 285? Reply with just the number.", temperature = 0)
dragon_compare(sft, rl)
```

`reward` in the comparison table is the mean total reward on held-out
prompts. Compare it with the same rewards computed on the starting model to
see what the stage bought: `dragon_evaluate()` on the starting run does not
know these rewards, so the quickest check is `dragon_generate()` on both with
`base = TRUE` on the RL run and a few prompts.

## Settings that matter

* **Reward variance is the signal.** If every sample in a group gets the
  same reward, that group contributes nothing. Prompts that are always
  solved or never solved waste samples; keep the data at a difficulty where
  the model sometimes succeeds.
* **`beta`** trades speed of change for stability. Raise it if replies
  degrade in ways the rewards do not see; lower it if nothing moves.
* **`temperature`** needs to be high enough that the group varies. 0.8 to
  1.2 is typical.
* **Learning rate** is small, 1e-6 to 2e-5. RL updates are noisier than
  supervised ones.
* **Length rewards** guard against the two failure modes of a length-blind
  reward: rambling until a right answer appears, or collapsing to a single
  token.

## Chaining with the other stages

Reinforcement learning is a stage like the others. A typical order is
fine-tune, then preference optimization for style, then RL for a
verifiable target, and each starts from the previous run. In a pipeline:

```{r}
dragon_pipeline("Qwen/Qwen2.5-0.5B-Instruct", list(
  dragon_step_train(tickets),
  dragon_step_reinforce(prompts, rewards = rewards, group_size = 6, max_new_tokens = 16),
  dragon_step_evaluate()
))
```
